---
title: "MEDICAL INSURANCE DATA ANALYSIS PROJECT"
author: "KILWAKE"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## SETTING WORKING DIRECTORY
```{r}
setwd("C:/Users/HomePC/Desktop/e-commerce project")
```

### DATA IMPORTATION
```{r}
insurance_data<-read.csv("insurance_data.csv")
head(insurance_data)
```
## structure
```{r}
str(insurance_data)
```



### DATA CLEANING

### missing values
```{r}
missing_value<-sum(is.na(insurance_data))
print(missing_value)
```


## removing duplicates
```{r}
duplicated_values<-sum(duplicated(insurance_data))
print(duplicated_values)
```


```{r}
library(dplyr)
insurance_data <- distinct(insurance_data)
```



```{r}
duplicated_values<-sum(duplicated(insurance_data))
print(duplicated_values)
```

```{r}
unique(insurance_data$sex)
```


### detecting outlier
```{r}
boxplot(insurance_data$charges,main="charges distribution")
boxplot(insurance_data$age, main = "age distribution")
boxplot(insurance_data$bmi,main="bmi distribution")
```



#### treating outliers

```{r}
Q1 <- quantile(insurance_data$charges, 0.25)
Q3 <- quantile(insurance_data$charges, 0.75)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

d_cleaned <-insurance_data[insurance_data$charges >=lower_bound &insurance_data$charges <=upper_bound,]
```


```{r}
Q1 <- quantile(d_cleaned$bmi, 0.25,na.rm = TRUE)
Q3 <- quantile(d_cleaned$bmi, 0.75,na.rm = TRUE)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

d_cleaned <-d_cleaned[d_cleaned$bmi>=lower_bound &d_cleaned$bmi<=upper_bound,]
```


```{r}
boxplot(d_cleaned$charges,main="charges distribution")
boxplot(d_cleaned$bmi,main="bmi distribution")

```

```{r}
Q1 <- quantile(d_cleaned$charges, 0.25,na.rm = TRUE)
Q3 <- quantile(d_cleaned$charges, 0.75,na.rm = TRUE)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

g_cleaned <-d_cleaned[d_cleaned$charges >=lower_bound &d_cleaned$charges <=upper_bound,]
```





```{r}
boxplot(g_cleaned$charges,main="charges distribution")
```


```{r}

Q1 <-quantile(g_cleaned$charges, 0.25,na.rm = TRUE)
Q3 <- quantile(g_cleaned$charges, 0.75,na.rm = TRUE)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

g_cleaned <-g_cleaned[g_cleaned$charges >=lower_bound &g_cleaned$charges <=upper_bound,]
```


```{r}
boxplot(g_cleaned$charges,main="charges distribution")
```


```{r}
Q1 <-quantile(g_cleaned$charges, 0.25,na.rm = TRUE)
Q3 <- quantile(g_cleaned$charges, 0.75,na.rm = TRUE)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

h_cleaned <-g_cleaned[g_cleaned$charges >=lower_bound &g_cleaned$charges <=upper_bound,]
```


```{r}
boxplot(h_cleaned$charges,main="charges distribution")
```


```{r}
unique(h_cleaned$children)
```

## DATA MANUPULATION
```{r}
 library(dplyr)

h_cleaned <- h_cleaned %>%
  mutate(age_group = case_when(
    age < 20 ~ "Young",
    age >= 20 & age <= 35 ~ "Youth",
    age >= 36 & age <= 59 ~ "Adult",
    age >= 60 ~ "Old"
  ))
head(h_cleaned)
```



```{r}
library(dplyr)

h_cleaned_processed1 <- h_cleaned %>%
  mutate(bmi_category = case_when(
    bmi < 18.5 ~ "Underweight",
    bmi >= 18.5 & bmi <= 24.9 ~ "Normal",
    bmi >= 25 & bmi <= 29.9 ~ "Overweight",
    bmi >= 30 ~ "Obese"
  ))
head(h_cleaned_processed1)

```

```{r}
h_cleaned_processed1 <- h_cleaned_processed1[!is.na(h_cleaned_processed1$bmi_category), ]
```

```{r}
h_filtered <- h_cleaned_processed1 %>%
  filter(smoker == "yes", age > 30) %>%
  arrange(desc(charges))

h_filtered
```

```{r}
h_cleaned_processed1 %>%
  group_by(region) %>%
  summarise(
    Min_Charges = min(charges, na.rm = TRUE),
    Max_Charges = max(charges, na.rm = TRUE)
  )
```



## DATA VISUALIZATION

```{r}
summary(h_cleaned_processed1)
```
The data set contains 1,104 observations on demographic characteristics, lifestyle factors, and insurance charges. Respondents range from 18 to 64 years, with an average age of 38 years. The gender distribution is relatively balanced, and non-smokers form the majority of the sample. Insurance charges vary widely, ranging from 1,122 to 23,568, indicating substantial differences in healthcare costs among individuals.






## Histogram
```{r}
library(ggplot2)

ggplot(h_cleaned_processed1, aes(x = charges)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Insurance Charges", x = "Charges", y = "Count",
   caption = "source: Insurance Dataset,2026")+
  
  theme_bw()
```

### Report of the histogram above

The distribution of insurance charges is right-skewed, with most charges concentrated at lower values and a long tail extending toward higher charges. This suggests that while the majority of individuals incur moderate healthcare expenses, a smaller proportion experience substantially higher medical costs. There also appears to be a small secondary concentration of charges at higher values, which may indicate the influence of specific risk factors such as smoking or obesity.


```{r}
library(ggplot2)

ggplot(h_cleaned_processed1, aes(x = bmi)) +
  geom_histogram(bins = 30, fill = "purple", color = "white") +
  labs(title = "Distribution of BMI", x = "BMI", y = "Number of respondents",
   caption = "source: Insurance Dataset,2026")+
  
  theme_bw()
```

###

The histogram shows the distribution of BMI among respondents in the insurance dataset. Most respondents have BMI values between 25 and 35, with the highest concentration around 30. This indicates that many individuals fall within the overweight category. Very low and very high BMI values are less common, showing that extreme BMI levels are rare in the dataset. Overall, the distribution appears approximately normal with moderate variation in BMI values.





```{r,fig.height=4,fig.width=4}
library(ggplot2)

ggplot(h_cleaned_processed1, aes(x = smoker, fill = smoker)) +
  geom_bar() +
  labs(title = "Bar chart showing smoking status", x = "Smoker", y = "Count", fill = "Smoker",
       caption = "Source:Insurance Data set 2026") +
  theme_minimal()
```

### Report of smokers

Non-smokers make the largest number in the data set




```{r}
h_cleaned_processed1 %>%
  group_by(age_group) %>%
  summarise(mean_charges = mean(charges)) %>%
  ggplot(aes(x = age_group, y = mean_charges, fill = age_group)) +
  geom_bar(stat = "identity", color = "black") +
  labs(x = "Age_group",
       y = "Mean Charges",
       title = "A bar plot showing mean insurance charges across age groups",
       caption = "source: Insurance Data set, 2026") +
  theme_bw()
```

### report of the bar plot

From the results old consumes more insurance(ages over 60 ) followed by adults(ages between 35 and 59),the young consumes less insurance




### pie chart of charges per region

```{r}
library(ggplot2)
library(dplyr)

h_cleaned_processed1 |>
  group_by(region) |>
  summarise(total_charges = sum(charges)) |>
  mutate(pct = total_charges / sum(total_charges),
         label = paste0(region, "\n", scales::percent(pct, accuracy = 0.1))) |>
  ggplot(aes(x = "", y = total_charges, fill = region)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 4) +
  labs(title = "Total Charges by Region", fill = "Region",caption = "Source:Insurance Data set,2026"
       ) +
  theme_void()
```
#### report of the pie  chart
The pie chart shows the northeast region accounts for the largest share of total charges at 27.0%, while the other three regions (northwest 24.5%, southwest 24.4%, southeast 24.0%) are fairly evenly splilt


### box plot of sex and charges

```{r}
h_cleaned_processed1|>
  ggplot(aes(x=sex,y=charges,fill = sex))+
  
  geom_boxplot()+
labs(x="sex",
       y="Charges",
       title = "A box plot of insurance charges and sex",
       )+
theme_bw()

```

### Report of insurance charges by sex

The boxplot compares insurance charges between male and female policyholders. Both groups display similar median charges, suggesting that sex on its own does not result in a large difference in typical insurance costs. However, the male group shows a slightly wider interquartile range, indicating greater variability in charges among males. Both groups contain high-end outliers, reflecting that a small number of individuals in each sex incur significantly higher medical costs. Overall, sex appears to have a limited effect on insurance charges in this dataset, and other variables such as smoking status or BMI are likely stronger drivers of cost.




```{r}
h_cleaned_processed1 |>
  ggplot(aes(x = bmi_category, y = charges)) +
  geom_boxplot(fill = "yellow") +
  labs(
    x = "BMI Category",
    y = " Charges",
    title = "Boxplot Showing Charges by BMI Category",
    caption = "Source: Insurance Dataset, 2026"
  ) +
  theme_classic()
  
```


### report

The boxplot shows that insurance charges are fairly consistent across BMI categories. Normal weight individuals have the highest median charges, followed closely by the Overweight and Obese groups, which are nearly equal in their central values. Underweight individuals tend to have the lowest charges. The wide spread across all categories suggests that BMI category alone is not a strong predictor of insurance costs.


### scatter plot of age and insurance

```{r}
h_cleaned_processed1 |>
  ggplot(aes(x = age, y = charges)) +
  geom_point(color = "purple") +
  labs(
    x = "Age",
    y = "Insurance Charges",
    title = "Scatterplot Showing Age and Insurance Charges",
    caption = "Source: Insurance Dataset, 2026"
  ) +
  theme_classic()
```

###

The scatterplot reveals a positive relationship between age and insurance charges — as age increases, charges tend to rise. Two distinct bands are visible: a lower band where charges increase gradually with age, and an upper band of consistently higher charges across all age groups, likely driven by high-risk individuals such as smokers. This pattern suggests that while age is a contributing factor, other characteristics significantly influence the level of charges.



```{r}
h_cleaned_processed1 |>
  ggplot(aes(x = bmi, y = charges)) +
  geom_point(color = "brown") +
  labs(
    x = "BMI",
    y = "Insurance Charges",
    title = "Scatterplot Showing BMI and Insurance Charges",
    caption = "Source: Insurance Dataset, 2026"
  ) +
  theme_classic()
```


###

The scatterplot shows no clear linear relationship between BMI and insurance charges. Points are widely scattered across all BMI values, suggesting that BMI on its own is a weak predictor of charges.



#### heat map
```{r}
library(corrplot)

h_cleaned_processed1 |>
  select(age, bmi, children, charges) |>
  cor(use = "complete.obs") |>
  corrplot(method = "color", addCoef.col = "black", tl.col = "black")
```
####

The correlation heatmap shows that age has the strongest relationship with insurance charges (r = 0.52), while BMI and number of children have negligible correlations with charges. Age is the only numeric variable with a meaningful linear association with insurance costs. Therefore, age should be considered as a key predictor when modeling insurance charges.




#### Hypothesis testing

##t-test
#H₀: Mean charges for males and females are equal.
#H₁: Mean charges for males and females are different.


```{r}
# T-test
t_test<-t.test(charges ~ sex, data = h_cleaned_processed1)
t_test
```
## conclusion
we reject the null hypothesis and conclude that the mean charges between sex is not equal to the mean charges of the other sex.

###Chi-square test

#H₀: Smoking and gender are independent.
#H₁: Smoking and gender are associated.
```{r}
table1 <- table(h_cleaned_processed1$sex,
                h_cleaned_processed1$smoker)

chisq.test(table1)
```

## conclusion

Since the p-value = 0.5739 more than the significance level of 0.05, we fail to reject the null hypothesis (H₀) and conclude that smoking and gender are independent



#ho:There is a significant association between gender and smoking status.
#h01: There is no significant association between gender and smoking status
#ho:There is a significant association between gender and smoking status.
#ho2:There is no significant association between gender and smoking status
#ho:There is significant association between region and smoking status.
#ho3:There isno significant association between region and smoking status.


```{r}
library(gtsummary)

h_cleaned_processed1 |>
  select(age, bmi, children, smoker, sex, region, charges) |>
  tbl_summary(by = smoker) |>
  add_ci() |>
  add_p(test = list(
    all_categorical() ~ "chisq.test",
    all_continuous() ~ "anova"
  ))
```

The chi-square test results indicate no significant association between smoking status and sex (p = 0.6), region (p = 0.5), or number of children (p > 0.9), since all p-values are greater than 0.05. However, smokers recorded substantially higher insurance charges compared to non-smokers.

## regression analysis

#H0: The predictors do not significantly influence insurance charges.
#H1: At least one predictor significantly influences insurance charges

```{r}
model_charges <- lm(charges ~ bmi+ age+children+smoker, data = h_cleaned_processed1)
summary(model_charges)
```

###Conclusion
The multiple linear regression results show that age, smoking status, and number of children are significant predictors of insurance charges since their p-values are less than 0.05. In particular, smokers incur substantially higher insurance charges compared to non-smokers. However, BMI is not a statistically significant predictor because its p-value (0.134) exceeds 0.05. The model has an R-squared value of 0.6897, indicating that approximately 68.97% of the variation in insurance charges is explained by the predictors included in the model.





